Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-17 manuscript/manuscript.tex
draft up to theory sec

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). In
general, however, statistics such as heterozygosity depend on the cumulative
autocorrelation factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. Under equilibrium levels of fitness
variation, the autocorrelation function for a neutral allele associated with
segment $i$ is the time-invariant function $C(t) = \[(1-r_i)(1-\kappa_i)\]^t$,
where $r_i$ is the recombination fraction to the segment and $\kappa_i$ is the
rate that the associated fitness variance decays due to selective dynamics.
Then, the cumulative autocorrelation is,

@@

The remaining pieces are expressions for the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. At this
point, we diverge from Santiago and Caballero (\citeyear{Santiago1998-bs},
\citeyear{Santiago2016-mu}) to note that the additive fitness variation is the
sum of additive \emph{genic} fitness variance $V_a$ (i.e. due to allele
frequencies) and the total linkage disequilibria $\delta_{LD}$ between selected
sites, $V_A = V_a + \delta_{LD}$. Considering only the genic variation, at
equilibrium $\Delta V_{a,i} = 0$; thus, the loss in genic fitness each
generation due to selective dynamics $-\kappa V_{A,i}$ must be equal to the
increase in fitness variation to new mutations $V_m$ each generation and
changes in the LD term \parencite{Bulmer1971-ae}. Note that the loss in genic
fitness due to drift is not time-invariant due to the build up of
autocorrelation, and considered elsewhere in the derivation (see Supplementary
Materials Equation \ref{eq:vardecay}). Similarly, the loss in variation due to
fixation of weakly deleterious occurs on much slower timescales and can be
ignored here. Then, $\kappa_i = \nicefrac{V_{m,i}}{V_{a,i}}$ (see Supplementary
Materials Equation \ref{eq:Z}).

@@

which is exactly the genome-wide model of background selection used in previous
studies \parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj} (note that
\cite{Santiago1998-bs} have shown a similar correspondence between their work
and BGS models with recombination from \cite{Nordborg1996-nq} and
\cite{Hudson1995-xc}).

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \parencite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent twice the rate of the ratchet of deleterious mutations in fixations
per generation, each fixation removes exactly the variation mutation puts in
$V_m = Us$, thus under a general equilibrium negative selection process,

\begin{align}
  V_{a} = (U - R)s. 
\end{align}

Our results differ from \parencite{Santiago2016-mu}, since we find from
simulations their expression models the additive \emph{genic} fitness
variation, not additive genetic fitness variation. Intuitively, the equilibrium
genic fitness variation is a balance between the rate of gain of genic fitness
variation due to mutation and its loss due to fixation. Note that since
mutation is the ultimate source of all mutations that fix or are lost, $U \ge
R$.

The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

@@

\subsection*{Effective Population Size Under Weak and Strong BGS}

>>>>

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). In
general, however, statistics such as heterozygosity depend on the cumulative
autocorrelation factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.
The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. Under equilibrium levels of fitness
variation, the autocorrelation function for a neutral allele associated with
segment $i$ is the time-invariant function $C(t) = [(1-r_i)(1-\kappa_i)]^t$,
where $r_i$ is the recombination fraction to the segment and $\kappa_i$ is the
rate that the associated fitness variance decays due to selective dynamics.
Then, the cumulative autocorrelation is,

@@

The remaining pieces are expressions for the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. At this
point, we diverge from Santiago and Caballero (\citeyear{Santiago1998-bs},
\citeyear{Santiago2016-mu}) to note that the additive fitness variation is the
sum of additive \emph{genic} fitness variance $V_a$ (i.e. due to allele
frequencies) and the total linkage disequilibria $\delta_{LD}$ between selected
sites, $V_A = V_a + \delta_{LD}$. Considering only the genic variation, at
equilibrium $\Delta V_{a,i} = 0$; thus, the loss in genic fitness each
generation due to selective dynamics $-\kappa V_{A,i}$ must be equal to the
increase in fitness variation to new mutations $V_m$ each generation and
changes in the LD term \parencite{Bulmer1971-ae}. Note that the loss in genic
fitness due to drift is not time-invariant due to the build up of
autocorrelation, and considered elsewhere in the derivation (see Supplementary
Materials Equation \ref{eq:vardecay}). Then, $\kappa_i =
\nicefrac{V_{m,i}}{V_{a,i}}$ (see Supplementary Materials Equation \ref{eq:Z}).

@@

which is exactly the genome-wide model of background selection used in previous
studies \parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}. Thus, the
classic background selection model is a special case of the more general theory
of Santiago and Caballero (\citeyear{Santiago2016-mu}), which they have shown
previously (\citeyear{Santiago1998-bs}).

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \parencite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent twice the rate of the ratchet of deleterious mutations in fixations
per generation, each fixation removes exactly the variation mutation puts in
$V_m = Us$ (we omit the segment index for clarity). Thus, under a general
equilibrium negative selection process,

\begin{align}
  \label{eq:Va}
  V_{a} = (U - R)s. 
\end{align}

Note that since mutation is the ultimate source of all mutations that fix or
are lost, $U \ge R$. Intuitively, this equilibrium genic fitness variation is a
balance between the rate of gain of genic fitness variation due to mutation and
its loss due to fixation. Our results differ from \parencite{Santiago2016-mu},
since we find from simulations their expression models the additive
\emph{genic} fitness variation, not additive genetic fitness variation (see
Section XXX), since it does not adequately capture linkage disequilibria
dynamics. Santiago and Caballero demonstrate Equation \eqref{eq:Va} through
Fisher's Fundamental Theorem of Natural Selection, but we find an alternative
proof (\cite{Higgs1995-xc}; see Supplementary Materials Section
\ref{supp:weak-strong}).

However, closed-form expressions for the rate of the ratchet $R$ have generally
been hard to find \parencite{Haigh1978-gt,Higgs1995-xc,Gessler1995-hz}. The key
insight of \parencite{Santiago2016-mu} is that the ratchet rate under draft is
given by the probability of fixation $p_f(N_e, s)$
\parencite{Kimura1962-su,Malecot1948-zv} with the draft-rescaled effective
population size, e.g. $R = N U p_f(N_e, s)$. Given this equation for the
ratchet and Equation \eqref{eq:polygenic_Ne} for $N_e$ under draft, we have a
system of non-linear equations that can be solved numerically for $N_e$ and
$R$.

\subsection*{Simulations of the Diploid Model}

@@

\subsection*{Effective Population Size Under Weak and Strong BGS} \label{supp:weak-strong}
